Understanding pandoc lua filters

### 2023-08-01 supplement/lua-filters/lua-filters.tex
Updated Lua filter supplement

\section{Introduction}
To understand pandoc filters, one must look at the document conversion process in
general. A document is read/parsed in one format to a common intermediate format.
In pandoc it is called an abstract syntax tree (AST). A document writer is designed 
to read the AST and write the contents in the form of desired article format.
For example converting markdown to HTML will involve these processes in pandoc.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.6\textwidth]{workflow.png}
  \caption{Conversion Workflow}
  \label{fig:workflow}
\end{figure}

The filters come into picture when there are certain elements which require 
customization. For example, suppose you are converting markdown to HTML and you
require the page to have automatic numbering for figures or tables. 
If you think about it in markdown there is no system of numbering figures/tables 
automatically, to retain 1:1 conversion the HTML output will miss the numbering as well. 
At this moment you would wish that there was some option to add such a feature,
but then given the infinite possibilities of customization it is not feasible to 
add each and every customization as an option in pandoc or the writer. 
To tackle this, the idea of filters arrived, wherein you could manipulate the AST
partly or fully and modify it before the writer could read the AST. Hence generating
an output you desire \citep{pandocfilters}.

\section{Writing lua filters}

@@

\begin{figure*}[!htbp]
\centering
\includegraphics[width=0.5\linewidth]{example.png}
\caption{Vanilla HTML5 output}
\label{fig:1}
\end{figure*}

\section{Using pandoc lua filters}

@@

Now we write a lua filter to manipulate the figures in \ref{code:3}
\subsection{TODO : explain lua filter in more detail.}
\begin{figure}[htbp]
\begin{verbatim}
figures = 0
is_fig = 0

@@

@@

\begin{figure}[htbp]
\centering
\includegraphics[width=0.5\linewidth]{example-filtered.png}
\caption{HTML5 output with desired filtering}
\label{fig:2}
\end{figure}

@@

\bibitem[Krewinkel, Lucero (2023)]{pandoc}
A.~ Krewinkel and A.~ Lucero,
\newblock pandoc 3.0 Release notes,
\newblock \emph{pandoc} \penalty0 2023
\newblock URL \url{https://pandoc.org/releases.html#pandoc-3.0-2023-01-18}

>>>>

\section{Introduction}
To fully understand pandoc filters, one must first grasp the document conversion process. A document in one format is read/parsed into an intermediate form. An abstract syntax tree (AST) is what pandoc calls it. A document writer is a specialized type-setter to read an AST and produces the contents in the chosen article format. Converting markdown to HTML, for example, will entail these pandoc operations. Figure\ref{fig:workflow} shows the conversion workflow withing pandoc.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.6\textwidth]{figures/workflow.png}
  \caption{Conversion Workflow}
  \label{fig:workflow}
\end{figure}

When specific elements require adjustment, the filters enter the picture. Assume you're converting markdown to HTML and want the page to have automated numbering for figures or tables. 
If you use markdown, there is no system for automatically numbering figures/tables; to retain 1:1 conversion, the HTML output will also lack numbering.

At this point, you might wish there was a way to add such a feature. Then there is the possibility of unlimited choices for customization, including every customization as an option in pandoc or the writer is not viable. 
To address this, the idea of filters emerged, in which you could change the AST partially and modify it before the writer could read it. As a result, you get the desired result \citep{pandocfilters}.

\section{Need of lua filters}

@@

\begin{figure*}[!htbp]
\centering
\includegraphics[width=0.5\linewidth]{figures/example.png}
\caption{Vanilla HTML5 output}
\label{fig:1}
\end{figure*}

@@

Now we write a lua filter to manipulate the figures in \ref{code:3}

\section{Writing pandoc lua filters}

\begin{figure}[htbp]
\begin{verbatim}
figures = 0
is_fig = 0

@@

Everything in Lua revolves around tables; for example, the pandoc AST or the document is one giant table with sub-tables.

In pandoc, there are two sorts of elements: \verb|Blocks| and \verb|Inlines|. \verb|Blocks| are complicated constructions constructed from simple pieces (\verb|Inlines|). A \verb|Para|, for example, is a \verb|Block| made up of several \verb|Str| \verb|Inlines|.

The first step in writing a filter is to select a target type, that is, which block or Inline this filter will target.
Following the selection, we name the function after the target type and each element in the argument as el.
For instance, in the case of our filter, it would be \verb|function Figure(el)|. Pandoc is smart enough to match the names of filter functions to the AST elements.

Now within the filter function we can assume that \verb|el| will contain the table object of a Figure element from the document. we can use many functions over it or \verb|Inlines| contained in it. One such function to walk over all the elements inside the block is \verb|walk_block|. walk functions are great to check or count the presence of certain elements in a block. we use \verb|walk_block| to check if there are any Image inline elements within the Figure block. This is because after pandoc 3 \citep{pandoc} Figure blocks can now contain elements other than Images as well.

If there is an Image then we append \verb|"Figure X :"| to the caption of the Figure element.
and return the modified element.

@@

\begin{figure}[htbp]
\centering
\includegraphics[width=0.5\linewidth]{figures/example-filtered.png}
\caption{HTML5 output with desired filtering}
\label{fig:2}
\end{figure}

@@

\bibitem[Krewinkel, Lucero (2023)]{pandoc}
A.~ Krewinkel and A.~ Lucero,
\newblock pandoc 3.0 Release notes,
\newblock \emph{pandoc} \penalty0 2023
\newblock URL \url{https://pandoc.org/releases.html}
